Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-21 paper/paper.tex
small tweaks to section 3, remove contractions

\paragraph{Regular Expressions}
Regular expressions can be encoded as types generated by base types,
$\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
grammar $\A$, we define the Kleene star $\A^{*}$ as a particular
\emph{inductive linear type} of linear lists, as shown in
\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining $\A^*$
to be a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
linearly consumes a parse of $\A$ and a parse of $\A^{*}$ and builds a
parse of $\A^{*}$. This linear consumption is defined by the linear
function type $\lto$. The linear function type $\A \lto \B$ defines functions that
take in parses of $\A$ as input, \emph{consume} the input, and return a parse of
$\B$ as output. The arrow, $\uparrow$, wrapping these
constructors means that the constructors are not consumed
upon usage, and so are \emph{non-linear} values themselves. That is, the
names $\nil$ and $\cons$ are function symbols that may be reused as many times
as we wish.

@@

Lastly, we include a type $\equalizer {a}{f}{g}$ that we call the
\emph{equalizer} of linear functions $f$ and $g$. We think of this
type as the ``subtype'' of elements of $A$ that satisfy the equation
$f\,a\equiv g\,a$. Note that it is important here that $f, g$
themselves are non-linearly used functions, as linear values cannot be
used in a type.  Equalizer types are not needed for non-linear types
since they can be constructed using the equality type as $\sum_{x:X}
f\,x=_Y g\,x$, but this construction can't be used for linear types
because it uses a \emph{dependent} version of the equality type, which
we cannot define as a linear type. While the equalizer type is not
used directly in defining any of our parsers or formal grammars, it is
used for several proofs, allowing for inductive arguments about our
indexed inductive types.

@@

Next, we introduce the most complex and important linear type
constructors of our development, \emph{indexed inductive linear
types}. We encode these by adding a mechanism for constructing initial
algebras of strictly positive functorial type expressions, following
prior work on inductive types
\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
\emph{strictly positive functorial} linear type expressions indexed by
a non-linear type $X$. We think of the elements of this type as
syntactic descriptions of linear types that are parameterized by
$X$-many variables standing for linear types that are only used in
strictly positive positions. Accordingly, the $\SPF\,X$ type supports
an operation $\el$ that interprets it as such a type constructor, as
well as an operator $\map$ that defines a functorial action on parse
transformers. The $\SPF\,X$ type supports constructors for a reference
$\Var~x$ to one of the linear type variables, a constant expression
that doesn't mention any type variables $K$, as well as tensor
products and additive conjunction and disjunction of type expressions.
Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
  }\else{the extended version of the paper}\fi that say that the
$\el$/$\map$ operations correspond to these descriptions of the
constructors.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\StringGram \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\StringGram \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.

>>>>

\paragraph{Regular Expressions}
Regular expressions can be encoded as types generated by base types,
$\oplus$, and $\otimes$, and the Kleene star $(\cdot)^{\ast}$.  For a
grammar $\A$, we define the Kleene star $\A^{*}$ as a particular
\emph{inductive linear type} of linear lists, as shown in
\cref{fig:kleenestarinductive}. Here $\A^{*} : \L$ means we are defining $\A^*$
to be a \emph{linear} type. $\A^{*}$ has two constructors: $\nil$, which
builds a parse of type $\A^{*}$ from nothing; and $\cons$, which
linearly consumes a parse of $\A$ and a parse of $\A^{*}$ and builds a
parse of $\A^{*}$. This linear consumption is defined by the linear
function type $\lto$. The linear function type $\A \lto \B$ defines functions that
take in parses of $\A$ as input, \emph{consume} the input, and return a parse of
$\B$ as output. The arrow, $\uparrow$, wrapping these
constructors means that the constructors are not consumed
upon usage, and so are \emph{non-linear} values themselves, which are not small. That is, the
names $\nil$ and $\cons$ are function symbols that may be reused as many times
as we wish.

@@

Lastly, we include a type $\equalizer {a}{f}{g}$ that we call the
\emph{equalizer} of linear functions $f$ and $g$. We think of this
type as the ``subtype'' of elements of $A$ that satisfy the equation
$f\,a\equiv g\,a$. Note that it is important here that $f, g$
themselves are non-linearly used functions, as linear values cannot be
used in a type.  Equalizer types are not needed for non-linear types
since they can be constructed using the equality type as $\sum_{x:X}
f\,x=_Y g\,x$, but this construction cannot be used for linear types
because it uses a \emph{dependent} version of the equality type, which
we cannot define as a linear type. While the equalizer type is not
used directly in defining any of our parsers or formal grammars, it is
used for several proofs, allowing for inductive arguments about our
indexed inductive types.

@@

Next, we introduce the most complex and important linear type
constructors of our development, \emph{indexed inductive linear
types}. We encode these by adding a mechanism for constructing initial
algebras of strictly positive functorial type expressions, following
prior work on inductive types
\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
\emph{strictly positive functorial} linear type expressions indexed by
a non-linear type $X$. We think of the elements of this type as
syntactic descriptions of linear types that are parameterized by
$X$-many variables standing for linear types that are only used in
strictly positive positions. Accordingly, the $\SPF\,X$ type supports
an operation $\el$ that interprets it as such a type constructor, as
well as an operator $\map$ that defines a functorial action on parse
transformers. The $\SPF\,X$ type supports constructors for a reference
$\Var~x$ to one of the linear type variables, a constant expression
that does not mention any type variables $K$, as well as tensor
products and additive conjunction and disjunction of type expressions.
Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
  }\else{the extended version\fi that say that the
$\el$/$\map$ operations correspond to these descriptions of the
constructors.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\StringGram \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\StringGram \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string does not
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.
